---
title: "Investigating the Factors Affecting Birthweight"
author: "Michael McCurtin, id = 21459584"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---


# Question of Interest 

Are the factors of smoking, previous history of hypertension or urinary irritability associated with whether babies were born with low birthweight (less than 2,500 grams)?

## Load the Libraries and Data Needed

The dataset you need is available in the `aplore3` package and we will also use the `tidyverse` package. These may not be installed already. If you need to install them then run following code in `R`. You will only need to do this once. After they have been installed you can just delete this chunk and text.

```{r}
install.packages(c("tidyverse", "aplore3"), 
                 repos = "https://cran.microsoft.com/", quiet = TRUE)
```

Load the required libraries so you can use them, and then make the birthweight data available ('lowbwt') as follows:   

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(aplore3)

data(lowbwt)
```

The low birthweight data is from the "Applied Logistic Regression" textbook by Hosmer and Lemeshow. The following is a description of the variables in this dataset.

|Name|	Description|
|:------|:------------------------------------------------------------------------|
|subject| identification code|
|low |low birthweight ("< 2500 g" or ">= 2500 g")|
|age |age of mother|
|lwt|weight at last menstrual period (pounds)|
|race |race (Black, White, Other)|
|smoke |smoked during pregnancy (Yes, No)|
|ptl |premature labour history (None, One, Two, etc.)|
|ht |history of hypertension (Yes, No)|
|ui |uterine irritability  (Yes, No)|
|ftv |number of visits to physician during 1st trimester (None, One, Two, etc.)|
|bwt |birthweight (in grams)|

## Subjective Impressions

The key variable of interest is `low` which represents whether a baby is born with low birthweight, defined as a birthweight below 2,500 grams.

```{r}
lowbwt %>% select(low) %>% table()
```

Let's explore the association between history of hypertension and low birthweight by tabulating the data.

```{r}
lowbwt %>% select(low, ht) %>% table()
```

It seems there were not many mothers with hypertension, but the proportions of low weight babies is very much higher for mothers suffering from hypertension status than those that were not.

```{r}
lowbwt %>% select(low, ht) %>% table() %>% prop.table(margin = 2)
```

Task: In the following `R` chunk explore the association between uterine irritability and whether the babies were born with low birthweight, using both the counts and appropriate percentages. Explain the results in words.

```{r}
lowbwt %>% select(low, ui) %>% table()
```


```{r}
lowbwt %>% select(low, ui) %>% table() %>% prop.table(margin = 2)

```

There were an equal amount of children in both weight classes born to mothers with with uterine irritability . However, the proportion of low-birthweight children from mothers suffering with uterine irritability is much higher.


Task: In the following `R` chunk explore the association between smoking status and whether the babies were born with low birthweight, using both the counts and appropriate percentages. Explain the results in words.

```{r}
lowbwt %>% select(low, smoke) %>% table()
```

```{r}
lowbwt %>% select(low, smoke) %>% table() %>% prop.table(margin = 2)

```

The data shows that it is much more likely for a low-birthweight child to have a smoking mother - the proportion of low-birthweight children with smoking mothers is much higher.

Now we will create some barcharts.

# Barchart of Low Birthweight

The following is a frequency plot of the low birthweight status.

```{r}
ggplot(lowbwt, aes(x = low, fill = low)) +
  geom_bar()
```

Task: In the following `R` chunk create a frequency plot of the smoking status.

```{r}
ggplot(lowbwt, aes(x = smoke, fill = smoke)) +
  geom_bar()
```


# Stacked Barchart of Low Birthweight by Hypertension Status

Below is a relative frequency plot of the low birthweight of the babies against the hypertension status of the mothers using a stacked barchart.

```{r}
ggplot(lowbwt, aes(x = ht)) +
     geom_bar(aes(fill = low), position = "fill") +
     ylab("Proportion")
```


Task: Create a stacked barchart of low birthweight by smoking status by inserting an `R` chunk and relevant code below.

```{r}
ggplot(lowbwt, aes(x = smoke)) +
     geom_bar(aes(fill = low), position = "fill") +
     ylab("Proportion")
```


Task: Create a stacked barchart of low birthweight by uterine irritability by inserting an `R` chunk and relevant code below.

```{r}
ggplot(lowbwt, aes(x = ui)) +
     geom_bar(aes(fill = low), position = "fill") +
     ylab("Proportion")
```

Task: Once you have created the plots, explain your interpretation of which factors are associated with low birthweight based on the three barcharts. State which factor you think is most associated with birthweight.


I believe that hypertension is most associated with birthweight - the proportion of low-birthweight children is roughly double when compared to mothers without hypertension. This doubling also holds true for uterine irritability but the proportion of low-birthweight children born to mothers with hypertension is higher.


The following `R` chunk produces a boxplot of the birthweight distribution.

```{r}
lowbwt %>% ggplot(aes(y = bwt)) +
  geom_boxplot() +
  labs(y = "Birthweight (in grams)")

```


Task: In the previous task you stated which factor you believe was most associated with birthweight, so you can explore the impact on the distribution in more detail. Create a graph of side-by-side boxplots comparing the birthweight distribution for each level of that factor (e.g. comparing mothers who had uterine irritability and those who did not), by inserting an `R` chunk and relevant code below.

[Hint: we used side-by-side boxplots in the week 4 lab and in the Exploratory Data Analysis worksheet]

```{r}
lowbwt %>% ggplot(aes(y = bwt, x = factor(ht))) +
  geom_boxplot() +
  labs(x = "Hypertension status of mother")

```



# Conclusion

Task: Write a short conclusion of whether you think low birthweight of babies can be predicted based on whether the mother smoked, has hypertension or uterine irritability.

I think a low birthweight can be predicted based on these factors. In all cases, the presence of these conditions results in a much higher proportion of babies with low birthweight when compared to their absence - more than a 100% increase in some cases such as hypertension.

Final Task: "knit" the file as a Word of PDF document and submit it via the relevant link on Blackboard before the deadline.



